A Mechanised Library for Generic (Co)Programming

### 2024-02-07 paper/paper.tex
Up to recursive hylos

A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
and a \emph{family of position types} indexed by shapes
(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
whose action on objects is given by the following definition:
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{Container} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape. The family of positions is
then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinine{Coq}{{p | Valid shape pos}} will be an OCaml
singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of shapes
and positions.,

\dcas{Solution to the previous problem}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (valid : Shape -> Pos -> bool) (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos = true} -> X }.
\end{minted}

\dcas{A shape, a position, and a decidable predicate on the validity of a
position in a shape define a functor}

\begin{minted}{Coq}
  Coercion is_true (b : bool) : Prop := b = true.
  Class functor (Shape Pos : Set) := { valid : Shape -> Pos -> bool }.
  Record Container `{functor Shape Pos} (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

@@

Reasoning about the equality of containers requires reasoning about the
equality of the contained elements. But the contained elements are given by a
function. This means that container equality needs to be extensional, if we
want to avoid the functional extensionality axiom. Furthermore, the equality of
functions also needs to be extensional. We capture this fact by associating
each type with an equivalence relation:

\begin{minted}{Coq}
  Class Equiv (T : Type) := 
  { equiv : T -> T -> Prop;
    e_refl : forall x, equiv x x;
    e_sym : forall x y, equiv x y -> equiv y x;
    e_trans : forall x y z, equiv x y -> equiv y z -> equiv x z;
  }.
\end{minted}

\dcas{Morphisms must respect equivalences}:

\begin{minted}{Coq}
  Record Morph `{Equiv A} `{Equiv B} := 
  { app : A -> B;
    app_ext : forall x y, equiv x y -> equiv (app x) (app y);
  }.

\dcas{We provide common morphisms for polynomials: sums, products, const,
  exponential}

\end{minted}

@@

\dcas{TODO: Hylo mechanisation, Hylo laws}

@@

@@

>>>>

A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
and a \emph{family of position types} indexed by shapes
(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
whose action on objects is given by the following definition:
\begin{minted}{Coq}
  Record App (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape. The family of positions is
then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
\begin{minted}{Coq}
  Record App (Shape : Set) (Pos : Set) 
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Container (Shape : Set) (Pos : Set) := {
   valid : Shape -> Pos -> bool
}.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
  Record App `{container Shape Pos} (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

\dcas{Examples of polynomials as container extensions?}

@@

Reasoning about the equality of containers requires reasoning about the
equality of the contained elements. But the contained elements are given by
function \mintinline{Coq}{contents}. This means that container equality needs
to be extensional, if we want to avoid the functional extensionality axiom.
Furthermore, the equality of functions also needs to be extensional. To deal
with the fact that different types will use different notions of equality, we
restrict to types with an associated equivalence relation, and \emph{proper
morphisms}, i.e. morphisms that respect the necessary equivalence relations.

\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).

Reserved Notation "x ~> y" (at level 95, right associativity, y at level 200).
Context `{eA : equiv A} `{eB : equiv B}.

 Structure morph :=
   MkMorph
     { app :> A -> B;
       app_eq : forall x y, x =e y -> app x =e app y
     }.
Arguments morph A {eA} B {eB}.
Notation "x ~> y" := (morph x y).
\end{minted}
\dcas{Simplify the above}

\dcas{We provide common morphisms for polynomials: sums, products, const,
exponential (Add examples?)}

By default, we associate every type with Coq's propositional equality, but allow
the use of other forms of equality. In most cases, we will use either propositional equality, extensional
equality, or a mixture of both, e.g. for container extensions, we use an arbitrary equality for shapes,
extensional equality for the contained elements, but Coq's propositional equality for positions:

\begin{minted}{Coq}
  Inductive AppR `{F : Container Sh P} (X : Type) {e : equiv X}
           (x y : App F X) : Prop :=
  | AppR_ext
      (Es : shape x =e shape y)
      (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
The reason for this is to avoid a circularity in the definitions: arbitrary
equivalences for positions would force us to define the contents as morphisms
from positions to the contained elements, but that would as well force us to
redefine container extensions so that they are only applicable to types with
arbitrary equalities and morphisms.  We will see in the following section why
this is problematic for defining initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

We associate some types with their standard equivalences:
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

@@

@@

Algebras on container extensions are defined as expected. The carrier of the
algebra is \mintinline{Coq}{A}, and the operations of the algebra are
morphisms of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}

There are \emph{two} initial algebras that are, as expected, unique up to
isomorphism. We show one such initial algebra here, and the second in the
following subsection, plus a proof that they are isomorphic:

\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This is why we could not define equality of container extensions generically:
the contained types may not have (yet) an associated equivalence. We could have
overcame this circularity by defining simultaneously LFix and its associated
equivalence relation. But that would complicate definitions unnecessarily.
\dcas{I know ... I need to rewrite/expand on this}

We provide a custom induction schema for initial algebras:

\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}

Equality of initial algebras of container extensions needs to be defined
recursively and extensionally:
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}

The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a 
regular recursive function:
\begin{minted}{Coq}
Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}

It is easy to show that this function is a proper morphism, and that it is
unique. We can use the fact that catamorphisms are proper morphisms to encode
them as a higher-order morphism.
\begin{minted}{Coq}
Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
  : forall x y, x =e y -> cata_f g x =e cata_f g y.
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...

Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{minted}{Coq}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{minted}

Their equivalence relation, and anamorphisms, and universal properties are
easy to encode: 
\begin{minted}{Coq}
  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}

With terminal coalgebras, we can encode another form of initial algebras. In
particular, the initial algebras based on terminal coalgebras are the carrier
types of terminal coalgebras, plus a proof that they are finite:

\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}

Using this predicate, we can define the notion of \emph{recursive coalgebras}
as those coalgebras that build finite trees. Recursive coalgebras are
implemented in Coq recursively on the proof that they produce finite trees.
Such proofs will be erased when extracting code to OCaml. This is standard
practice when building general recursive functions in Coq\footnote{\dcas{This
technique had a name in the community, but I cannot remember. Something
predicates}}.
\begin{minted}{Coq}
Notation rana_f__ h :=
  (fix f x H :=
    let hx := h x in
    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).

Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

Lemma rana_univ A {eA : equiv A} (h : RCoalg A) (f : A ~> LFix F)
    : f =e rana h <-> f =e l_in \o fmap f \o h.
\end{minted}

\subsection{Recursive Hylos}

Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
\begin{minted}{Coq}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
: forall (x : A), RecF c x -> B
:= fix f x H :=
  match c x as h0
        return
        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
        B
  with
  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
  end (RecF_inv H).
\end{minted}

\begin{minted}{Coq}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{minted}

Recursive hylo universal property allows us to prove common optimisations:

\begin{minted}{Coq}
    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
    : f2 \o hylo g1 h1 =e hylo g2 h1.
  Proof.
    rewrite hylo_univ.
    rewrite fmap_comp.
    rewrite compA.
    rewrite <- E2.
    rewrite <- compA.
    rewrite <- compA.
    rewrite (compA g1).
    rewrite <- hylo_unr.
    reflexivity.
  Qed.

  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

  Corollary cata_ana_hylo `(F : Container Sh P) `{equiv A} `{equiv B}
    (g : Alg F B) (h : RCoalg F A)
  : cata g \o rana h =e hylo g h.
\end{minted}
